Cross Validation Approaches for Penalized Cox Regression

### 2018-05-17 main.tex
Edits to introduction

\section{Introduction}
Cox propotional hazard regression, since its original proposal in 1972, is by far the most common regression approach for analyzing survival data. Cox regression is built upon partial likelihood. Under the propotional hazard assumption, partial likelihood still preserves most properties of a full likelihood, while the estimation of the baseline can be skipped. Despite its popularity, Cox model is limited by this semiparametric nature: it is challenging to evaluate the model's predictive accuracy.

Suppose there are n subjects and m observed failure events. Use i for indexing observations, j for indexing events sorted by their failure time. For the i th subject, its covarites will be denoted by $X_{i}$; $\delta_{i}$ denotes whether an event is observed or not for subject i. $t_{i}$ denotes the time observed for subject i; $t_{j}$ denotes the failure time for the j th event.  In the Cox model \citep{Cox1975}, the hazard function for an observation i is given by 
\begin{equation}h_{i}(t) = h_{0}(t) \text{exp}( X_{i}^{T} \beta)\end{equation} 
where $h_{0}$ is the baseline hazard and $e^{X_{i}^{T} \beta}$ is the relative risk. The estimation of the coefficients $\beta$s is obtained by maximizing the partial likelihood. For each individual subject, its contribution to the partial likelihood is 
\begin{equation} l_{i}(\beta) = \left \{\frac{\text{exp} ( X_{i}^{T} \beta)}{\sum_{ k \in R(t_{i})}\text{exp} ( X_{k}^{T} \beta)}\right \}^{\delta_{i}} \end{equation} \note{introduce notation $l_{i}$} , where $R(t_{i})$ denotes the set of observations at risk at time $t_{i}$.

For all n subjects, the partial likelihood is constructed as a product of all individual contributed partial likelihood:
\begin{equation} L(\beta) =\prod_{i = 1}^{n} l_{i}(\beta) = \prod_{j=1}^{m} \frac{\text{exp} ( X_{j}^{T} \beta)}{\sum_{ k \in R(t_{j})}\text{exp} ( X_{k}^{T} \beta)}\end{equation}. As a semi-parametric model, Cox regression only gives estimations of the $\beta$ coefficients, without estimating the baseline hazard. Hence the interpretation of the Cox model is only valid in a relative sense.

Penalized Cox Regression is an extension of the regular cox regression model. Estimates of the $\beta$ coefficients are obtained by minimizing the objective function \begin{equation}  Q(\beta |X, y) = - \frac{1}{n}  \sum_{j=1}^{m} \text{log} \left \{\frac{\text{exp} ( X_{j}^{T} \beta)}{\sum_{ k \in R(t_{j})}\text{exp} ( X_{k}^{T} \beta)}) \right \} + P_{\lambda}(\beta). \end{equation} \citep{Simon2011}. The first part of the objective function is based on the partial likelihood. The second part is a penalty term that depends on some regularization parameters. In this paper, only the LASSO penalty $P_{\lambda}(\beta) = \lambda (\sum_{j} |\beta_{j}|)$ was considered. The LASSO cox regression can be used to build prediction models for survival outcome based on high dimensional data, where the number of covariates p $\gg$ n. An example of its application would be to make predictions of cancer patients' the overall survival time with their certain gene expressions.

@@

One common approach for selecting the tuning parameter $\lambda$ for linear regression and logistic regression is via the K-fold cross validation. The data set would be split into K folds. One fold would be treated as the test set and the other K - 1 folds as the training set. The model would first be built on the training set, then fitted to the test set to obtain cross validation error (CVE).  CVE would be calculated for each candidate $\lambda$, then the one that minimizes cross validated error would be selected.

However, for penalized cox regression, it is challenging to carry out cross validation. Since the partial likelihood of the cox model does not give estimates for the baseline, it is not meaningful to compare the coefficients if they are not estimated from different models and different observations. Only if the baseline is the same, then the coefficients can be interpreted as relative risks. This is against the idea of cross validation, which splits the data and builds models on different subsets.

%\begin{figure}
 %   \centering
	%	\includegraphics[height= 4cm ]{./figures/cv.png}	
    %\caption{K-fold Cross Validation}
%\end{figure}

%ZZZZZZZZZZZZZ
\par Currently, there are two common cross-validation approaches for Cox regression, both having its disadvantages. The most intuitive approach to conduct cross validation in Cox regression is i) fit the model to training set then ii) calculate cross-validated partial likelihood based on the observations from the test set as a measure for the model's predictive accuracy. As will be discussed in Section 2, this approach is unstable when number of events gets small. The second approach was proposed by Verweij et al. \citep{Verweij1993}. This is a stabilized version of cross-validated log likelihood, which has been widely used and implemented in R packages such as $\tt{glmnet}$ \citep{glmnet}. Although this approach fixes the stability issue, it tends to yield more conservative models in practice. We hope to improve cross-validation procedure for Cox regression to achieve a good balance between stability and performance.

\par In this paper, we proposed two alternative ways to carry out cross validation. Instead of cross-validating over partial likelihood, we propose to cross validate over the linear predictors of the regression model and cross validate over the deviance residual \citep{Therneau1990}of the Cox model. We conducted simulation studies to compare how those methods perform in both low dimensional and high dimensional settings with LASSO penalty. We compared the two proposed methods with the existing cross-validated partial likelihood approaches. We showed that the linear predictor approach outperforms other cross validation approaches in various scenarios. We found out that using cross validation to select tuning parameters for LASSO penalized cox regression tends to be a conservative approach in general. Finally, we apply those methods to two high dimensional data sets from real studies with time-to-event outcomes.

  \subsection{Cross Validated Likelihood} 
Suppose a data set of n observations is split into K folds and the ith fold is left out, then a partial likelihood can be built over the K -1 folds and yields maximum likelihood estimates of the coefficients, denoted by $\hat{\beta}_{-i}$. An intuitive way to carry out cross validation is to use the partial likelihood on the test data as the cross validated error: \begin{equation}cvl = \sum_{i=1}^{K} l_{i}(\hat{\beta}_{-i})\end{equation} This is implemented in the $\tt{glmnet}$ package as the $\tt{ungrouped}$ option. We refer to this definition as standard cross-validated partial likelihood.

\par An alternative way to calculate cross-validated log likelihood was proposed by Verweij and Van Houwelingen. They defined cross-validated log likelihood of leaving the ith fold out to be \begin{equation}cvl = \sum_{i=1}^{K} \{l(\hat{\beta}_{-i}) - l_{-i}(\hat{\beta}_{-i})\}. \end{equation} When the ith fold is left out, $l(\hat{\beta}_{-i})$ is the log partial likelihood evaluated at $\hat{\beta}_{-i}$ with all k folds of observations. $l_{-i}(\hat{\beta}_{-i})$ is the log partial likelihood evaluated at $\hat{\beta}_{-i}$ with observations from the other K-1 folds. We refer to this definition as Verweij and Van Houwelingen's cross-validated partial likelihood. This is implemented in the $\tt{glmnet}$ package as the default $\tt{grouped}$ option.

>>>>

\section{Introduction}
Since its original proposal \citep{Cox1972}, Cox proportional hazards regression has become the most common regression approach for analyzing survival data.  Cox regression utilizes a partial likelihood construction under an assumption of proportional hazards to estimate the regression coefficients without having to specify the underlying baseline hazard.  The ability to avoid choosing a specific parametric distribution for the survival time is very attractive, as time-to-event data is often poorly described by fully parametric models, but this semiparametric approach comes at a cost.

The Cox regression approach can estimate relative risks, but without estimating the baseline hazard, it cannot make any predictions concerning the absolute failure time for any given individual.  This poses a challenge to assessing the predictive accuracy of a given Cox regression model.  The challenge is particularly relevant for penalized Cox regression \citep{Tibshirani1997,Fan2002}, as the assessment of predictive accuracy via cross-validation is the standard method for selecting the regularization parameter and deciding upon a model.

Standard cross-validation involves of dividing the data into $K$ folds, then fitting the model on $K-1$ of those folds (the training set) and assessing prediction accuracy on the remaining fold (the testing set).  For Cox regression, the estimated coefficients allow us to quantify the risk for each subject in the training set relative to other members of the training set, but it is not obvious how to use those estimates to quantify the model's accuracy in the testing set.

One approach is to simply calculate the partial likelihood based on the observations in the test set as a measure for the model's predictive accuracy.  In doing so, we calculate the risk for each member of the test set relative to other members of the test set.  One drawback to this approach is that it becomes unstable when the size of the the test set is small.  In particular, it cannot be applied to leave-one-out cross-validation (LOOCV), as we must have at least two observations in the test set to compare their risk relative to each other.

%In this paper, we briefly review existing approaches for meeting this challenge, then propose two new methods for carrying out cross-validation in Cox regression models.

%As a semi-parametric model, Cox regression only gives estimations of the $\beta$ coefficients, without estimating the baseline hazard. Hence the interpretation of the Cox model is only valid in a relative sense.

% NOTE: I don't know that we need this paragraph; cross-validation is pretty common.
%One common approach for selecting the tuning parameter $\lambda$ for linear regression and logistic regression is via the K-fold cross validation. The data set would be split into K folds. One fold would be treated as the test set and the other K - 1 folds as the training set. The model would first be built on the training set, then fitted to the test set to obtain cross validation error (CVE).  CVE would be calculated for each candidate $\lambda$, then the one that minimizes cross validated error would be selected.

To overcome this drawback, an alternative approach was proposed by \citet{Verweij1993}.  Their approach, which we describe in detail in Section~\ref{Sec:cox-cv-existing}, stabilizes the cross-validated log likelihood, enabling its use even when the number of subjects in each fold is small.  This approach has been widely used and is implemented in R packages such as {\tt glmnet} \citep{glmnet}.  Although this approach fixes the stability issue, we demonstrate here that in practice, it tends to behave conservatively in terms of model selection.

%We hope to improve cross-validation procedure for Cox regression to achieve a good balance between stability and performance.

In this paper, we propose two alternative ways to carry out cross validation. Instead of cross-validating over partial likelihood, we propose to cross validate over the linear predictors of the regression model and cross validate over the deviance residual \citep{Therneau1990}of the Cox model. We conducted simulation studies to compare how those methods perform in both low dimensional and high dimensional settings with LASSO penalty. We compared the two proposed methods with the existing cross-validated partial likelihood approaches. We showed that the linear predictor approach outperforms other cross validation approaches in various scenarios. We found out that using cross validation to select tuning parameters for LASSO penalized cox regression tends to be a conservative approach in general. Finally, we apply those methods to two high dimensional data sets from real studies with time-to-event outcomes.

\note{Move this paragraph to Section 2?} In the Cox model, the hazard function for subject $i$ is given by 
\begin{equation}
  h_{i}(t) = h_{0}(t) \text{exp}( X_{i}^{T} \beta),
\end{equation} 
where $h_{0}$ is the baseline hazard and $e^{X_i^{T} \beta}$ is the risk for subject $i$ with covariates $X_i$ relative to that baseline.  The estimation of the coefficients is obtained by maximizing a partial likelihood.  Letting $t_i$ denote the time on study for subject $i$ and $\delta_{i}$ indicate whether or not an event is observed for subject $i$, each subject's individual contribution to that partial likelihood is
\begin{equation}
  l_{i}(\beta) = \left \{\frac{\text{exp} ( X_{i}^{T} \beta)}{\sum_{ k \in R(t_{i})}\text{exp} ( X_{k}^{T} \beta)}\right \}^{\delta_{i}},
\end{equation}
where $R(t_{i})$ denotes the set of observations at risk at time $t_{i}$.  The partial likelihood for the entire sample of $n$ subjects is then given by:
\begin{equation}
  L(\beta) =\prod_{i = 1}^{n} l_{i}(\beta).
\end{equation}
%% \begin{equation}
%%   L(\beta) =\prod_{i = 1}^{n} l_{i}(\beta) = \prod_{j=1}^{m} \frac{\text{exp} ( X_{j}^{T} \beta)}{\sum_{ k \in R(t_{j})}\text{exp} ( X_{k}^{T} \beta)}.
%% \end{equation}

Penalized Cox Regression is an extension of the regular cox regression model. Estimates of the $\beta$ coefficients are obtained by minimizing the objective function
\begin{equation}
  Q(\beta |X, y) = - \frac{1}{n}  \sum_{j=1}^{m} \text{log} \left \{\frac{\text{exp} ( X_{j}^{T} \beta)}{\sum_{ k \in R(t_{j})}\text{exp} ( X_{k}^{T} \beta)}) \right \} + P_{\lambda}(\beta).
\end{equation}
The first part of the objective function is based on the partial likelihood. The second part is a penalty term that depends on some regularization parameters. In this paper, only the LASSO penalty $P_{\lambda}(\beta) = \lambda (\sum_{j} |\beta_{j}|)$ was considered. The LASSO cox regression can be used to build prediction models for survival outcome based on high dimensional data, where the number of covariates p $\gg$ n. An example of its application would be to make predictions of cancer patients' the overall survival time with their certain gene expressions.

@@

\subsection{Cross Validated Likelihood} 
\label{Sec:cox-cv-existing}

Suppose a data set of n observations is split into K folds and the ith fold is left out, then a partial likelihood can be built over the K -1 folds and yields maximum likelihood estimates of the coefficients, denoted by $\hat{\beta}_{-i}$. An intuitive way to carry out cross validation is to use the partial likelihood on the test data as the cross validated error: \begin{equation}cvl = \sum_{i=1}^{K} l_{i}(\hat{\beta}_{-i})\end{equation} This is implemented in the $\tt{glmnet}$ package as the $\tt{ungrouped}$ option. We refer to this definition as standard cross-validated partial likelihood.

An alternative way to calculate cross-validated log likelihood was proposed by Verweij and Van Houwelingen. They defined cross-validated log likelihood of leaving the ith fold out to be \begin{equation}cvl = \sum_{i=1}^{K} \{l(\hat{\beta}_{-i}) - l_{-i}(\hat{\beta}_{-i})\}. \end{equation} When the ith fold is left out, $l(\hat{\beta}_{-i})$ is the log partial likelihood evaluated at $\hat{\beta}_{-i}$ with all k folds of observations. $l_{-i}(\hat{\beta}_{-i})$ is the log partial likelihood evaluated at $\hat{\beta}_{-i}$ with observations from the other K-1 folds. We refer to this definition as Verweij and Van Houwelingen's cross-validated partial likelihood. This is implemented in the $\tt{glmnet}$ package as the default $\tt{grouped}$ option.
